Exploring the Soil Microbiome coupling with the Crete island Data Cube

### 2024-04-11 manuscript.tex
add appendix table

This study uses the ISD Crete data that have been deposited
in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB21776.
The raw sequences (fastq files) and metadata (xml files) were downloaded with custom scripts using the ENA API \cite{Yuan2023}.
Specific information regarding the ISD sampling, DNA extractions, PCR and sequencing
are discussed here \cite{holm2024}.

@@

The complilation of Crete data cube has multiple spatial data layers of global,
european, greek or cretan scale. 
The Copernicus CORINE Land Cover has 3 layers resolution that classify the land
use and cover in shapefile format \cite{CLC2023}. 
WorldClim 2.0 contains global climatic data for 12 variables, e.g annual mean
temperature and annual precipitation \cite{Fick2017}.
We also utilised teh Environmentally Sensitive Areas Index to desertification (ESAI) 
of Greece dataset \cite{KARAMESOUTI2018266}. Additionally we included the 
Global Aridity Index and Potential Evapotranspiration Database \cite{zomer2022version}.
Crete geological bedrock formations were downloaded from IGME.
Handling and analysis of these data was done with the sf and stars R packages \cite{Pebesma2023}.
%World Database on Protected Areas (WDPA) \cite{Hanson2022}
%Average NDVI, 2000-2020
%Harmonized World Soil Database
\subsection{Network inferrence}\label{network_inferrence}

The network inferrence was facilitated with FlashWeave 0.19.2 \cite{Tackmann2019}
in Julia language version 1.9.3 \cite{Julia-2017}. To use FlashWeave we reduced 
the abandunce table of the ASVs to keep the ones at the genus or species level.
In addition, we filtered these taxa that appeared in ### samples and had more than 
#### mean relative abundance. The subsequent network analyses were carried out
with the igraph R package \cite{Csardi2006}.

\subsection{Tools}\label{tools}
%PEMA for OTU inferrence \cite{Zafeiropoulos2020}
%U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}

For taxa function annotation we used the manually curated FAPROTAX database and python script \cite{Louca2016}.
Whereas for the differential abundance analysis with ANCOM-BC2 R package \cite{Lin2023}.
Numerical ecology analyses,e.g diversity indices, NMDS, PERMANOVA we calculated
with the vegan R package \cite{vegan}.
For PCoA ordination we used ape R packege \cite{Paradis2004} and UMAP python library\cite{mcinnes2018umap-software}.
Visualisation was implemented with ggplot2 \cite{ggplot22016} and pheatmap \cite{Kolde2019}.
The environment we worked had Python 3.11.4 and R version 4.3.2 \cite{rcoreteam}. 
Finally, computations were performed on HPC infrastrure of HCMR \cite{Zafeiropoulos2021}.

@@

@@

Future works, shotgun metagenomics, long reads, need for even higher resolution samplings.

@@

%\begin{appendices}
%
%\section{Appendix 1}\label{appendix1}
%
%Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
%pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
%vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
%a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
%urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris.
%
%
%\section{Appendix 2}\label{appendix2}
%\subsubsection{Subsubsection title of first appendix}\label{subsubsec3}
%
%Fusce mauris. Vestibulum luctus nibh at lectus. Sed bibendum, nulla a faucibus semper, leo velit ultricies tellus, ac
%venenatis arcu wisi vel nisl. Vestibulum diam. Aliquam pellentesque, augue quis sagittis posuere, turpis lacus congue
%quam, in hendrerit risus eros eget felis.
%
%\end{appendices}
\section{Data and Code}
The documentation and scripts developed for this study are available in
\href{https://github.com/savvas-paragkamian/crete_soil_microbiome/}{Crete soil microbiome github repository}.
This repository contains all the necessary scripts for the data retrieval,
filtering and ASV inferrence, taxonomy assignment, data integration of spatial data, 
functional annotation and the subsequent analyses and visualisation.
Code is structured to be reproducible and interoperable.

>>>>

This study uses the ISD Crete data that have been deposited
in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB21776.
The raw sequences (fastq files) and metadata (xml files) were downloaded with custom scripts using the ENA API \cite{Yuan2023}.
Specific information regarding the ISD sampling, DNA extractions, PCR and sequencing
protocols are discussed here \cite{holm2024}.

@@

The complilation of Crete data cube has multiple spatial data layers of global,
european, greek or cretan scale. 
The Copernicus CORINE Land Cover has 3 layers resolution that classify the land
use and cover in shapefile format \cite{CLC2023}. 
WorldClim 2.0 contains global climatic data for 12 variables, e.g annual mean
temperature and annual precipitation \cite{Fick2017}.
We also utilised teh Environmentally Sensitive Areas Index to desertification (ESAI) 
of Greece dataset \cite{KARAMESOUTI2018266}. Additionally we included the 
Global Aridity Index and Potential Evapotranspiration Database \cite{zomer2022version}.
Geological formations shapefiles were downloaded from the geoportal of
Decentralized Administration of Crete, which were developed by
Crinno-Emeric Group project\footnote{\url{https://geoportal.apdkritis.gov.gr/gis/apps/storymaps/stories/19690f65abbe4e8ab0141b2fe7261a8c}}.
Handling and analysis of these data was done with the sf and terra R packages \cite{Pebesma2023}.
%World Database on Protected Areas (WDPA) \cite{Hanson2022}
%Harmonized World Soil Database

\subsection{Integrative Analysis and Annotations}\label{int_analysis}
The network inferrence was facilitated with FlashWeave 0.19.2 \cite{Tackmann2019}.
To use FlashWeave we reduced 
the abandunce table of the ASVs to keep the ones at the genus or species level.
In addition, we filtered these taxa that appeared in SOSO samples and had more than 
SOSO mean relative abundance. The subsequent network analyses were carried out
with the igraph R package \cite{Csardi2006}.

For taxa function annotation we used the manually curated FAPROTAX database and python script \cite{Louca2016}.
Whereas for the differential abundance analysis with ANCOM-BC2 R package \cite{Lin2023}.
Numerical ecology analyses,e.g diversity indices, NMDS, PERMANOVA we calculated
with the vegan R package \cite{vegan}.
For PCoA ordination we used ape R packege \cite{Paradis2004} and UMAP python library\cite{mcinnes2018umap-software}.

\subsection{Tools}\label{Coding environment}
%PEMA for OTU inferrence \cite{Zafeiropoulos2020}
%U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}

Visualisation was implemented with ggplot2 \cite{ggplot22016} and pheatmap \cite{Kolde2019}.
The environment we worked had Python 3.11.4, R version 4.3.2 \cite{rcoreteam}
and Julia language version 1.9.3 \cite{Julia-2017}in Julia language version 1.9.3 \cite{Julia-2017}.
Finally, computations were performed on HPC infrastrure of HCMR \cite{Zafeiropoulos2021}.

@@

Crete's ecosystems are mostly semi-arid, whereas in the mountain ranges there 
are areas classified as dry sub-humid and humid.

Parent material and infuences on soil functions and bacterial communities.

@@

The pillar of data integration is the unrestricted open data across disciplines and 
the open source software.
Deciphering and validating the results presented here requires future work.
Shotgun metagenomics and metatranscriptomics can uleash the functional potential of
topsoil along with other advancements like long reads sequencing. Higher resolution
samplings using grid system will enhance the resolution and also the resampling of
ISD sites in different time points will provide additional insights to the complex soil 
functions.

@@

\begin{appendices}

\section{Appendix 1}\label{appendix1}

Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris.

\begin{table}[]
    \caption{Taxonomic depth of ASVs and the unique number of taxa of each level.\label{table:asv_taxonomy}}%
\begin{tabular}{@{}lll@{}}
classification depth & Total ASV    & Total taxa \\
Kingdom              & 1974         & 2          \\
Phylum               & 4034         & 33         \\
Order                & 38517        & 193        \\
Class                & 24157        & 83         \\
Family               & 71355        & 287        \\
Genus                & 90137        & 1166       \\
Species              & 9120         & 1338       \\
Total                & $\sim$239000 & 3102      
\end{tabular}
\label{table:asv_taxonomy}
\end{table}

%\section{Appendix 2}\label{appendix2}
%\subsubsection{Subsubsection title of first appendix}\label{subsubsec3}
%
%Fusce mauris. Vestibulum luctus nibh at lectus. Sed bibendum, nulla a faucibus semper, leo velit ultricies tellus, ac
%venenatis arcu wisi vel nisl. Vestibulum diam. Aliquam pellentesque, augue quis sagittis posuere, turpis lacus congue
%quam, in hendrerit risus eros eget felis.
%
\end{appendices}
\section{Data and Code}
The documentation and scripts developed for this study are available in
\href{https://github.com/savvas-paragkamian/crete_soil_microbiome/}{Crete soil microbiome github repository}.
This repository contains all the necessary scripts for the data retrieval,
filtering and ASV inferrence, taxonomy assignment, data integration of spatial data, 
functional annotation and the subsequent analyses and visualisation.
Code is structured to be reproducible and interoperable.
